
\thispagestyle{plain}			% Suppress header
\section*{Acknowledgements}
Countless people have influenced my journey as I study and implement methods to learn more about complexity science. 
\medskip

I am extremely grateful for the opportunity to have worked at Kaleido Biosciences this year under the guidance of my industrial advisor and manager Stephan Reiling. Because of his mentorship my scientific and computational thinking has grown considerably. At the same time he and the members on the IDS team have helped show me the relevancy of computational methods in the life science and biotech industries. 

\medskip
At Chalmers I would like to thank Professor Rebecka J\"{o}rnsten for working with me as my academic advisor and thesis examiner. Her statistical learning class helped show how much more there is to explore in the life sciences with the use of novel computational methods. Her insight helped me to choose this portion of my microbiome research to present as a thesis. 
I would also like to thank my thesis opponent, Dimitrios Karypidis, for his friendship and assistance with challenging my thesis defense.
% Thank: Stephan, Kaleido, Rebecka Jörnsten, family, and friends

\medskip
None of this would have been possible without my parents, T.J. Dufresne and T.A. Rosko, who have always done all that they could to foster my thirst for knowledge about the world. It's because of them and their constant support that I have followed this path to explore physics and complexity science. My sister, Becca Jane, has been there through the ups and downs, and I am so thankful to have her by my side offering encouragement and enthusiasm. My grandparents, Dick and Jane, who continuously share their love and belief in me and have helped make this opportunity to study at Chalmers a reality. My grandfather, Stan, who has cheered me on and given me advice as I begin my career. 

\medskip
Finally, I'd like to thank all of my friends that have shared their time with me while I've been in this program. My friends from MPCAS, \"{O}vre Fogelbergsgatan 3, Tjugosex, The Hostel, Bates College, Summit, and especially those who have helped me with proofreading and editing. 

\medskip

The work presented in this thesis represents only a portion of the research that I worked on this year. I hope that you find it informative, and are excited about future work in human microbiome research. 

\vspace{1.5cm}
\hfill
Wade Rosko, Boston/Gothenburg, June 2019

\newpage				% Create empty back of side
\thispagestyle{empty}
\mbox{}